---
title: "Estimation of Rate Constant Parameters for Acid/Base Attack of Aspirin"
author: "Jim Hughes"
date: "15 April 2019"
output:
  word_document: default
---

_Note: This document as well as all code in relation to this document can be found
at https://github.com/jhhughes256/acidbase_fit _

### Aim
The aim of this document is to provide an example of estimating rate constant
parameters for the acid/base attack of a molecule. In this document the example
is aspirin.

### Software
Throughout this document, the R programming software for statistical analysis
will be used. This software is free and open source and is the gold standard
for statistical analysis. Here we will fit a non-linear model to rate constant
and pH data.

R lets you define objects, which contain either single numbers, strings of 
numbers, words, lists or tables of data. These can then be used with functions, 
a set of actions that will be performed on an object to give an output.

For example the statement `min(x)`, uses the `min` function on `x` an object in
the environment. The `min` function finds the minimum value, so for a string
of numbers from 0 to 20, this function will return a value of 0.

If further understanding is required, please refer to tutorials online.

### Acid/Base Attack Equation
First the reactions taking place should be considered. For aspirin, there are 
six potential reactions. Acid, base and water attack for the unionised form
of aspirin and acid, base and water attack for the ionisied form. However,
only four reactions at most can be observed.

To determine how many reactions can be observed, the log of the rate constant
`log k` is plotted against `pH`. To do this in R:

```{r rawplot}
# First load a graphical package
  library(ggplot2)  # graphical package

# Set the ggplot2 theme to your preference
  theme_bw2 <- theme_set(theme_bw(base_size = 14))
  theme_update(plot.title = element_text(hjust = 0.5))
  
# Read in the data to be plotted
# This data is the observed data consisting of matched logk and pH values
  aspirin_data <- read.csv("D:/2019 PostDoc/Des/aspirin_hydrolysis.csv")
  
# Then plot said data
# The plot is built up over multiple lines, adding elements to the overall plot
  p <- NULL
  p <- ggplot(data = aspirin_data)
  p <- p + geom_point(aes(x = pH, y = logk), shape = 1, colour = "blue")
  p <- p + labs(x = "pH", y = "log k (min^-1)")
  p
```

From this we can see four distinct reactions. The acid and water catalysed
degradation of unionised aspirin, and the base and water catalysed degradation
of ionisied aspirin. Therefore the equation for the rate constant is:

$$
k = k_H[H^+]\frac{[H^+]}{[H^++K_a]} + k_{H_2O}\frac{[H^+]}{[H^++K_a]} +
    k'_{H_2O}\frac{[H^+]}{[H^++K_a]} + k'_{OH}\frac{K_w}{[H^+]}\frac{K_a}{[H^++K_a]}
$$

The equation has four rate constants to be estimated: $k_H$, $k_{H_2O}$, 
$k'_{H_2O}$ and $k'_{OH}$.

### Estimating Rate Constants
To estimate the rate constants we will use R to define a function, which 
calculates the total rate constant $k$ and is dependent on what estimated
rate constants are provided as input.

```{r pred}
# First the aspirin data would be read in, however this is already done, so it
# is commented out
  # aspirin_data <- read.csv("D:/2019 PostDoc/Des/aspirin_hydrolysis.csv")
  
# Now we define our function, which is called k_pred
# It has par and pH as inputs
# par is a string of numbers describing the estimated rate constants
# pH is a string of numbers describing the pH that we wish to calculate k at
  k_pred <- function(par, pH) {
  # Define objects with fixed values
  # These include the values for Kw and Ka
    Kw <- 1E-14
    pKa <- 3.57
    Ka <- 10^-pKa
    
  # Determine values dependent on pH
    H <- 10^-pH
    fAH <- H/(H+Ka)
    fA <- Ka/(H+Ka)
    
  # Define parameter values to be used in the equations
  # Square brackets are used to extract the 1st, 2nd, 3rd and 4th number from par
    k1H <- par[1]
    k1H20 <- par[2]
    k2H20 <- par[3]
    k2OH <- par[4]
    
  # Determine each part of the rate constant equation and put them in a table
    data.frame(
      pH = pH,
      k1H = k1H*H*fAH,  # acid-catalysed unionised
      k1H2O = k1H20*fAH,  # water-catalysed unionised
      k2H2O = k2H20*fA,  # water-catalysed ionised
      k2OH = k2OH*(Kw/H)*fA,  # base-catalysed ionised
  # Finally determine the total rate constant
      k = k1H*H*fAH + k1H20*fAH + k2H20*fA + k2OH*(Kw/H)*fA
    )
  }
```

Using the new function `k_pred` we can now calculate $k$ given we have values
for $k_H$, $k_{H_2O}$, $k'_{H_2O}$ and $k'_{OH}$ as an object and a string of
values for pH that we wish to predict $k$ at.

Next we must create a function that determines how accurate a given set of
rate constants are at determining `k`. This is done by calculating an objective
function, which is a value describing the likelihood that the given rate
constants are correct.

```{r estim}
# This function as two inputs: par and data
# par is the same as the input seen in the k_pred function
# It is a string of parameters describing the rate constants to be estimated
# It also includes an additional parameter, which is the error parameter
# This is located at the end of the string
# data is the observed data consisting of matched logk and pH values
  estimate_fn <- function(par, data) {
  # Define independent variable, named x for the x axis on a plot
    x <- aspirin_data$pH
    
  # Define dependent variable, named y for the y axis
  # the log k is converted to k: k = 10^{log k}
    y <- 10^aspirin_data$logk  
    
  # Define predicted dependent variable
  # This prediction is based on the values of par
  # We use the k_pred function we defined to calculate this predicted value for k
  # $k is added to the end of the function as we only want the k term, as 
  #   k_pred also provides a number of other values
  # It is named yhat, as the hat symbol denotes a prediction
    yhat <- k_pred(par, x)$k
    
  # Test (log) likelihood that prediction is correct
  # This function `dnorm` is a density function and determines how normally
  #   distributed the observed values are around the predicted values.
  # The normal distribution has a mean of the predicted value, and a standard
  #   deviation dependent on the value of the error parameter multiplied by
  #   the predicted value. This is known as proportional error.
    loglik <- dnorm(y, mean = yhat, sd = par[5]*yhat, log = T)
    
  # Return objective function value to be minimised
    return(-1*sum(loglik))
  }
```

So now when use our new `estimate_fn` function, we provide it with `par`, which
is a string of numbers, with the initial numbers being the rate constants that
wish to be estimated (in this case: $k_H$, $k_{H_2O}$, $k'_{H_2O}$ and 
$k'_{OH}$). The final number is the proportional error term (in this case: 
`par[5]`) and is used when determining the objective function.

The last step of estimation is to use a "hill climbing" algorithm to find the
best set of rate constants that describe our observed data. When given a set
of initial estimates for a set of parameters, this algorithm will run a function
repeatedly using those parameters, changing the values slightly each time it 
runs the function. It aims to climb to the highest point, that is, determine
the objective function value that is the best fit for the observed data.

```{r optim}
# First we define our rate constants (in days^-1)
# These are found from literature
  k_constants <- c(2.5, 0.01, 0.1, 5000)

# Now we define our initial parameters
# the rate constants are converted to min^-1
# also an initial estimate for error is defined, and added to the end of the 
# our parameters
  err <- 0.1
  init_par <- c(k_constants/24/60, err)
  init_par

# Then the hill-climbing algorithm function `optim` is used to determine our
# optimised rate constants, which will be stored in `k_opt`
# We must provide optim with:
# - the initial parameter values (par = init_par)
# - the function that calculates the objective function value (fn = estim_fn)
# - the preferred hill-climbing algorithm (method = "L-BFGS-B")
  k_opt <- optim(
    par = init_par, 
    fn = estimate_fn, 
    method = "L-BFGS-B",
    lower = c(init_par[1:4]*0.8, 0.001), 
    upper = c(init_par[1:4]*1.2, 1)
  )
  # L-BFGS-B is used as you can give an lower and upper bounds to the search
  #   for the optimal parameter values. lower = ... and upper = ... are provided
  #   with separate lower and upper bounds for each parameter.
  # Here we are using 80% of the initial estimate for the rate constants as the
  #   lower bound, and 120% as the upper bound.
  # The error parameter is given a lower bound of 0.001 and an upper bound of 1.
  #   Therefore an lower bound of 0.1% error and an upper bound of 100% error.
  
# Finall we need to extract the optimised parameters from our k_opt object
# We use the `head` function to get the first values, minus 1 value from the end
#   as we are not interested in the error parameter
# In doing so we convert them from min^-1 back to days^-1
  opt_par <- head(k_opt$par, -1)*60*24
  k_constants
  opt_par
```

Finally we can re-calculate the total rate constant $k$ with our optimised 
rate constants ($k_H$, $k_{H_2O}$, $k'_{H_2O}$ and $k'_{OH}$) and plot them
against our observed data.

``` {r plot}
# Using our k_pred function we create our predicted data
# For the pH we provide k_pred with a sequence of numbers from 0 to 14, with
#   the gaps between values being 0.1 (e.g. 0, 0.1, 0.2, ... 13.9, 14)
  pred_data <- k_pred(par = opt_par, pH = seq(0, 14, by = 0.1))

# None of these values are in the log domain, so we must convert them all
  pred_data$logk <- log10(pred_data$k)
  pred_data$logk1H <- log10(pred_data$k1H)
  pred_data$logk1H2O <- log10(pred_data$k1H2O)
  pred_data$logk2H2O <- log10(pred_data$k2H2O)
  pred_data$logk2OH <- log10(pred_data$k2OH)
  
# We also need to convert the observed data to be in the same units as our
#   predicted data (days^-1)
  aspirin_data$k_min <- 10^aspirin_data$logk
  aspirin_data$k_day <- 60*24*aspirin_data$k_min
  aspirin_data$logk_day <- log10(aspirin_data$k_day)
  
# Create plot
# The observed data is plotted as points
# The predicted rate constant is plotted as a solid line
# Each of the components of the rate constant is plotted as a dashed line
# coord_cartesian sets the limits for each axis
  p <- NULL
  p <- ggplot(data = pred_data)
  p <- p + geom_point(aes(x = pH, y = logk_day), data = aspirin_data, 
    shape = 1, colour = "blue")
  p <- p + geom_line(aes(x = pH, y = logk), size = 1)
  p <- p + geom_line(aes(x = pH, y = logk1H), linetype = "dashed")
  p <- p + geom_line(aes(x = pH, y = logk1H2O), linetype = "dashed")
  p <- p + geom_line(aes(x = pH, y = logk2H2O), linetype = "dashed")
  p <- p + geom_line(aes(x = pH, y = logk2OH), linetype = "dashed")
  p <- p + ylab("log k (d^-1)")
  p <- p + coord_cartesian(xlim = c(0.5, 11.5), ylim = c(-4, 2.5))
  p

```
